Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2015-12-21 paper.tex
approaching submission ready

\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$,
Nick~Higginbottom$^1$, Knox~S.~Long$^2$, Stuart~A.~Sim$^3$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

@@

\begin{abstract} 
Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing quasar spectra to 
synthetic spectra of simple biconical disc wind models, produced with our state-of-the-art 
Monte Carlo radiative transfer and photoionization code PYTHON.  
Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars 
for plausible wind models, but only if the X-ray luminosity was limited to about  
10$^{43}$~erg~s$^{-1}$. Here, we investigate the  degree to which
clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame ultraviolet (UV) 
that have the characteristics of quasars with X-ray luminosities as 
large as 10$^{45}$ erg s$^{-1}$. 
We conclude, perhaps not surprisingly, that clumping does maintain the ionization 
state of the wind necessary for strong BAL features in the rest-frame UV at more 
realistic X-ray luminosities. We examine the X-ray properties of our fiducial clumped model
and find good agreement with existing X-ray samples of AGN and quasars.
In addition, the dense, X-ray heated wind produces strong recombination and collisionally 
excited line emission in, e.g., \la\ and \civline, to emerge at the low inclination, 
`Type 1 quasar-like' angles. At the highest inclinations, the synthetic spectra 
possess prominent \mg\ and \al\ BALs, the absorption features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce all the emission lines seen in
quasar spectra with the correct equivalent width 
ratios in any of the models in our initial simulation grid,
and we find an angular dependence of emission line equivalent width despite the 
remarkable similarities in the observed spectral properties of BAL and non-BAL quasars.
Overall, our work suggests that a biconical wind model
reproduces much of the qualitative behaviour expected from a unified model. 
However, we cannot yet provide quantitative matches with quasar properties at all 
viewing angles; whether disc winds can successfully unify quasars is therefore 
still an open question.
\end{abstract}

% \begin{keywords}
% galaxies: active; accretion, accretion discs; (galaxies:) quasars: emission lines;
% (galaxies:) quasars: absorption lines
% \end{keywords}

@@

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind. 
According to this paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows  in radio-loud (RL) sources \citep{zhou2006,ghoshpunsly2007}.

@@

Despite the clear importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma, meaning that it is difficult to prevent 
the wind becoming over-ionized and `failing' in the presence of strong X-rays. 
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
An additional or alternative solution is that the wind is clumped 
\citep[e.g.][]{hamann2013}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma while still maintaining the same mass-loss 
rate and column density.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
Clumping can be caused by magnetic confinement \citep{dekool1995},
or the instabilities inherent to line-driven winds 
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984,owockirybicki1985}.
Additionally, clumping is required to explain the electron scattering wings of emission lines formed
in line-driven hot star winds \citep{hillier1991eswingsmodel}. Complex substructures 
are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to line-driven instabilities 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
Nevertheless, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario.

We have been engaged in a project to determine whether it is possible to simulate the properties of the spectra of AGN, including BALQSOs, using simple kinematic prescriptions for biconical disc winds using a Monte Carlo radiative transfer (MCRT) code that calculates the ionization structure of the wind and simulates the spectra from such a system \citep[][hereafter H13 and H14]{simlong2008,sim2010,higginbottom2013,H14}.  The results have been encouraging in the sense that in H13, we showed we could produce simulated spectra that resembled that of BALQSOs, as long as the luminosity of the X-ray source was relatively low, of order \POW{43}{\LUM} and the mass loss rate was relatively high, of order the mass accretion rate.  However, at higher X-ray luminosities, the wind was so ionized that UV absorption lines were not produced.  In addition, and in part due to limitations in our radiative transfer code, the model failed to produce spectra with strong emission lines at any inclination angle.

Here we attempt to address both of these issues, by introducing clumping into our model and a more complete treatment of H and He into our radiative transfer calculations.   The remainder of this paper is organized as follows:
In section 2, we describe some of the important photoionization 
and MCRT aspects of the code. We then outline the model in section 3, including 
a description of our clumping implementation and success criteria. 
Section 4 contains the results from a clumped model, which we discuss results
in comparison to observational data. Finally, we summarise our findings in section 5.

@@

The code operates as follows:   This outflow is discretized into $n_x \times n_z$ cells in a 2.5D
cylindrical geometry with azimuthal symmetry. From some initial conditions in each cell, the code first calculates the ionization structure of the wind in a series of iterations.  Each iteration in an ``ionization cycle'' consists of generating photons, actually photon packets, 
from an accretion disc and central object, and calculating how these photon 
bundles scatter through the wind (eventually escaping the outflow or hitting the disk). 
The ionization and temperature structure is updated based on the properties of the 
radiation field in each cell, and the process is repeated. The radiative transfer and thermal
balance is carried out according to the principle of radiative equilibrium, 
which assumes both statistical equilibrium and that radiative heating balances radiative cooling.
Once the ionization structure has converged, it is held fixed, 
and synthetic spectra are generated at specific inclination 
angles in a series of "spectral cycles". LK02 provide a more detailed  
description of the original code; various improvements have been made 
since then and are described by \cite{simmacro2005}, H13 and M15.  
We focus here on the specific changes made for this study
intended to improve the ionization calculation of H and He
and to allow for clumping in the wind.

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}). 
which are reprocessed by `macro-atoms' with associated transition probabilities.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as a macro-atom, because 
we expect recombination to be important
in determining their level populations and resultant line emission, 
and because we are especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-ions. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and, in principle, allows 
the direct addition of extra physical processes that would previously have necessitated 
approximate treatments.

@@

We include two sources of r-packets in our model:
An accretion disc and central X-ray source.
The accretion disc is assumed to be geometrically thin, but optically thick.
Accordingly, we treat the disc as an ensemble of blackbodies with a 
\cite{shakurasunyaev1973} effective temperature profile. 
The emergent SED is then determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are opaque, meaning
that r-packets that strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~r_G$, where 
$r_G = GM_{BH}/c^2$ is the gravitational radius.
For a $10^9~M_\odot$ BH, this is equal to $8.8\times10^{14}~{\rm cm}$ 
or $\sim10^{-4}~{\rm pc}$.

The X-ray source is treated as an isotropic sphere at the ISCO,
which emits r-packets according to a power law in flux with index $\alpha_X$, of the form
\begin{equation}
F_X (\nu) = K_X \nu^{\alpha_X}.
\end{equation}
The normalisation, $K_X$ of this power law is such that it 
produces the specified 2-10~keV luminosity, $L_X$.
In addition to the disc and X-ray source, 
the wind is able to reprocess radiation. However, new 
photon packets are not produced in the wind (as in LK02). 
Instead, this reprocessing is dealt with by enforcing strict
radiative equilibrium ({\em modulo} adiabatic cooling; see section~2.3)
via an indivisible energy packet
constraint (see Lucy 2002, M15).  
% {\bf I confess, I do not understand what is being said here.  I would like to understand it better, and I would like it explained better.  How do we avoid, generating for example recombination photons from non-macro atoms.}

@@

In our model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010)
\item Line emission emerging at low inclinations, with $EW\sim40$\AA in \civline\ \citep[e.g. ][]{shen2011}
\item H recombination lines with $EW\sim50$\AA in \la\ \citep[e.g. ][]{shen2011}
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000}_{3000} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/link8.png}
\caption
{
Contour plots showing the logarithm of some important 
physical properties of the outflow. Symbols are defined in the text.
The solid black line marks a sphere at $1000~r_G$.
The dotted lines show the $72^\circ$ and $78^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
The line luminosities, $L$, represent the luminosity of photons
escaping the Sobolev region for each line. These photons do not
necessarily escape to infinity.
}
\label{fig:wind}
\end{figure} %fullpage

\noindent
Figure~\ref{fig:wind} shows the physical properties of the wind.
The wind rises slowly from the disc at first, with clump densities 
of $n_H \sim 10^{11}~\rm{cm^{-3}}$ close to the disc plane, 
where $n_H$ is the local number density of H.
The flow then accelerates over a scale length of $R_V=10^{19}~\rm{cm}$
up to a terminal velocity equal to the escape velocity at the streamline base
($\sim10,000~\rm{km~s^{-1}}$). This gradual acceleration results in
a wind that exhibits a stratified ionization structure, with low ionization material
in the base of the wind giving way to highly ionized plasma further out.
This is illustrated in figure~\ref{fig:wind} 
by the panels showing the ion fraction $F=n_j/n_{tot}$ of some important ions.
With a clumped wind, we are able to produce the range of ionization states observed
in quasars and BALQSOs, while adopting a realistic $2-10$ keV X-ray luminosity
of $L_{X}=10^{45}~\rm{erg~s^{-1}}$. Without clumping, this wind would be over-ionized 
to the extent that opacities in e.g., \civ\ would be entirely negligible (see H13).

@@

The total line luminosity also increases dramatically compared to H13. 
This is primarily due to the higher X-ray luminosity increasing the bound-free
heating in the plasma, which avoids over-ionization due to clumping.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
% This is equivalent to $\beta_{ul} n_u A_{ul}$, where the three quantities
% represent the Sobolev escape probability, upper level number density and 
% Einstein A coefficient for the line. 
As shown in figure~\ref{fig:wind},
the \civline\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civline\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec_new.png}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled to $1350$\AA
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure} %fullpage

@@

The main motivation for including a treatment of clumping was
to avoid over-ionization of the wind in the presence of strong X-rays. 
Having verified that strong BALs appear in the synthetic spectra,
it is also important to assess whether the X-ray properties of this
fiducial model agree well with quasar and BALQSO samples for the relevant
inclinations.

>>>>

\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$, 
Knox~S.~Long$^2$, Stuart~A.~Sim$^3$, Nick~Higginbottom$^1$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

@@

\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to about 
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow maintains the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent width ratios, and we find an angular dependence of emission line equivalent width despite the remarkable similarities in the observed spectral properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

\begin{keywords}
galaxies: active; accretion, accretion discs; (galaxies:) quasars: emission lines;
(galaxies:) quasars: absorption lines; methods: numerical; radiative transfer.
\end{keywords}

@@

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, knigge2008, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind. 
According to this paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies suggest that the wind is roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows  in radio-loud (RL) sources \citep{zhou2006,ghoshpunsly2007}.

@@

Despite the importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains uncertain. 
Several driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma, meaning that it is difficult to prevent 
the wind becoming over-ionized and `failing' in the presence of strong X-rays. 
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
An additional or alternative solution is that the wind is clumped 
\citep[e.g.][]{hamann2013}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma while still maintaining the same mass-loss 
rate and column density.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
Clumping can be caused by magnetic confinement \citep{dekool1995},
or the instabilities inherent to line-driven winds 
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984,owockirybicki1985}.
Additionally, clumping is required to explain the electron scattering wings of emission lines formed
in line-driven hot star winds \citep{hillier1991eswingsmodel}. Complex substructures 
are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to line-driven instabilities 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
In summary, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario.

We have been engaged in a project to determine whether it is possible to simulate the properties of 
the spectra of AGN, including BALQSOs, using simple kinematic prescriptions for biconical disc winds.
To address this question, we use a Monte Carlo radiative transfer (MCRT) code that calculates the ionization structure of the wind 
and simulates the spectra from such a system 
\citep[][hereafter H13 and H14]{simlong2008,sim2010,higginbottom2013,H14}.  The results have been encouraging in the sense that in H13, we showed we could produce simulated spectra that resembled that of BALQSOs, as long as the luminosity of the X-ray source was relatively low, of order \POW{43}{\LUM} and the mass loss rate was relatively high, of order the mass accretion rate.  However, at higher X-ray luminosities, the wind was so ionized that UV absorption lines were not produced.  In addition, and in part due to limitations in our radiative transfer code, the model failed to produce spectra with strong emission lines at any inclination angle.

Here we attempt to address both of these issues, by introducing clumping into our model and a more complete treatment of H and He into our radiative transfer calculations.   
The remainder of this paper is organized as follows:
In section 2, we describe some of the important photoionization 
and MCRT aspects of the code. We then outline the model in section 3, including 
a description of our clumping implementation and success criteria. 
Section 4 contains the results from a clumped model, 
with comparisons to observational data, as well as some discussion. 
Finally, we summarise our findings in section 5.

@@

The code operates as follows:   This outflow is discretized into $n_x \times n_z$ cells in a 2.5D
cylindrical geometry with azimuthal symmetry. 
From some initial conditions in each cell (typically $T_e \sim 40,000$K in AGN, with 
dilute LTE abundances), 
the code first calculates the ionization structure of the wind in a series of iterations. 
Each iteration in an ``ionization cycle'' consists of generating photons, actually photon packets, 
from an accretion disc and central object, and calculating how these photon 
bundles scatter through the wind (eventually escaping the outflow or hitting the disk). 
The ionization and temperature structure is updated based on the properties of the 
radiation field in each cell, and the process is repeated. The radiative transfer and thermal
balance is carried out according to the principle of radiative equilibrium, 
which assumes both statistical equilibrium and that radiative heating balances radiative cooling.
Once the ionization structure has converged, it is held fixed, 
and synthetic spectra are generated at specific inclination 
angles in a series of "spectral cycles". LK02 provide a more detailed  
description of the original code; various improvements have been made 
since then and are described by \cite{simmacro2005}, H13 and M15.  
We focus here on the specific changes made for this study
intended to improve the ionization calculation of H and He
and to allow for clumping in the wind.

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}),
which are reprocessed by `macro-atoms' with associated transition probabilities.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as macro-atoms, because the process is computationally
intensive, and we expect recombination to be particularly important
in determining their level populations and resultant line emission.
We are also especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-ions. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and
prepares for a more accurate treatment of other physical processes in future. 
% in principle, allows 
% the direct addition of extra physical processes that would previously have necessitated 
% approximate treatments.

@@

We include two sources of r-packets in our model:
An accretion disc and a central X-ray source.
The accretion disc is assumed to be geometrically thin, but optically thick.
Accordingly, we treat the disc as an ensemble of blackbodies with a 
\cite{shakurasunyaev1973} effective temperature profile. 
The emergent SED is then determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are opaque, meaning
that r-packets that strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~r_G$, where 
$r_G = GM_{BH}/c^2$ is the gravitational radius.
For a $10^9~M_\odot$ BH, this is equal to $8.8\times10^{14}~{\rm cm}$ 
or $\sim10^{-4}~{\rm pc}$.

The X-ray source is treated as an isotropic sphere at the ISCO,
which emits $r$-packets according to a power law in flux with index $\alpha_X$, of the form
\begin{equation}
F_X (\nu) = K_X \nu^{\alpha_X}.
\end{equation}
The normalisation, $K_X$ of this power law is such that it 
produces the specified 2-10~keV luminosity, $L_X$.
Photons, or $r$-packets, produced by the accretion disc and central X-ray source
are reprocessed by the wind. This reprocessing is dealt with by enforcing strict
radiative equilibrium ({\em modulo} adiabatic cooling; see section~2.3)
via an indivisible energy packet constraint (see Lucy 2002, M15).  
% In addition to the disc and X-ray source, 
% the wind is able to reprocess radiation. However, new 
% photon packets are not produced in the wind (as in LK02). 
% Instead, this reprocessing is dealt with by enforcing strict
% radiative equilibrium ({\em modulo} adiabatic cooling; see section~2.3)
% via an indivisible energy packet
% constraint (see Lucy 2002, M15).  
% {\bf I confess, I do not understand what is being said here.  I would like to understand it better, and I would like it explained better.  How do we avoid, generating for example recombination photons from non-macro atoms.}

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010)
\item Line emission emerging at low inclinations, with $EW\sim40$\AA in \civline\ \citep[e.g. ][]{shen2011}
\item H recombination lines with $EW\sim50$\AA\ in \la\ \citep[e.g. ][]{shen2011}
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000}_{3000} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/link8.png}
\caption
{
Contour plots showing the logarithm of some important 
physical properties of the outflow. The spatial scales are
logarithmic and the $x$ and $z$ scales are not the same.
Symbols are defined in the text.
The solid black line marks a sphere at $1000~r_G$.
The dotted lines show the $72^\circ$ and $78^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
The line luminosities, $L$, represent the luminosity of photons
escaping the Sobolev region for each line. These photons do not
necessarily escape to infinity.
}
\label{fig:wind}
\end{figure} %fullpage

\noindent
Figure~\ref{fig:wind} shows the physical properties of the wind.
The wind rises slowly from the disc at first, with densities within clumps
of $n_H \sim 10^{11}~\rm{cm^{-3}}$ close to the disc plane, 
where $n_H$ is the local number density of H.
The flow then accelerates over a scale length of $R_V=10^{19}~\rm{cm}$
up to a terminal velocity equal to the escape velocity at the streamline base
($\sim10,000~\rm{km~s^{-1}}$). This gradual acceleration results in
a wind that exhibits a stratified ionization structure, with low ionization material
in the base of the wind giving way to highly ionized plasma further out.
This is illustrated in figure~\ref{fig:wind} 
by the panels showing the ion fraction $F=n_j/n_{tot}$ of some important ions.
With a clumped wind, we are able to produce the range of ionization states observed
in quasars and BALQSOs, while adopting a realistic $2-10$ keV X-ray luminosity
of $L_{X}=10^{45}~\rm{erg~s^{-1}}$. Without clumping, this wind would be over-ionized 
to the extent that opacities in e.g., \civ\ would be entirely negligible (see H13).

@@

The total line luminosity also increases dramatically compared to the unclumped model
described by H13. This is because the denser outflow can absorb the increased
X-ray luminosity without becoming over-ionized, leading to a hot plasma which
produces strong collisionally excited line emission.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
% This is equivalent to $\beta_{ul} n_u A_{ul}$, where the three quantities
% represent the Sobolev escape probability, upper level number density and 
% Einstein A coefficient for the line. 
As shown in figure~\ref{fig:wind},
the \civline\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civline\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec_new.png}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled to $1350$\AA\
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure} %fullpage

@@

The main motivation for adding clumping to the model was
to avoid over-ionization of the wind in the presence of strong X-rays. 
Having verified that strong BALs appear in the synthetic spectra,
it is also important to assess whether the X-ray properties of this
fiducial model agree well with quasar and BALQSO samples for the relevant
inclinations.
